 

### 2023-03-11 frontmatter.Rmd
Revisions throughout and final version for External Reader

First and foremost, I would like to thank the members of my dissertation committee, Stéphane Verguet, Jessica Cohen, Karestan Koenen, and Margaret McConnell. I thank the committee chair, Stéphane Verguet, for his service, and for his exceptional mentorship as my academic advisor. Throughout the last four years, he has consistently offered unwavering support and insightful guidance throughout my academic journey, and in the moments where I faltered or doubted myself most, his confidence and cheer always helped see me through.

Jessica Cohen has been a constant source of joy and inspiration throughout my studies, and her exceptional teaching and scholarship have set a standard to aspire to. In addition to being a brilliant scholar, she is an incredible mentor and instructor, bringing genuine joy to the rigorous study of challenging questions. Her enthusiasm for econometrics and public health is infectious, and I feel extremely grateful to have learned so much by her example.

@@

I am also deeply grateful to my coauthors, without whom the research in these pages would be immeasurably lessened, and to the reviewers whose thoughtful reflections strengthened every page. I thank Shekhar Saxena for his guiding leadership in global mental health and for welcoming me into the field when I first arrived at Harvard, and I thank Christopher Golden for welcoming me into his lab and providing every opportunity, guidance, and support a student could wish for throughout the development of the second paper of this thesis.

@@

The community in the Department of Global Health and Population has been a treasured home away from home, and I am grateful to so many of its members for the opportunity to learn at their side. My deep appreciation goes to David Bloom, David Canning, Margaret Kruk, and Vikram Patel, whose classes and coffee chats all have shaped my interests and the course of my career.

Among the acknowledgements of our department's dissertations past, two people who have tirelessly worked to support students on their journeys are highlighted, without fail. This dissertation is no exception: to Barbara Heil and Allison Conary, I owe a debt of gratitude for all the mountains they have moved and the uphill struggles they have seen me soundly through.

I also would like to thank the many other individuals at the School of Public Health who have made the last four years so memorable and full of warm memories. I am thankful to Claudette Agustin, Matthew Boccuzzi, Jarvis Chen, and Bruce Villineau for all the care and welcome they have shared with me and with each cohort of students in the Population Health Sciences program.

@@

Across the Charles River are many dear colleagues who have dedicated themselves to ensuring that Harvard feel like home to all its students. To Jackie Yun and Janet Daniels at the GSAS Student Center, and to Sheila Thomas, Katie Saibara, and Karina Gonzalez Herrera of the GSAS Office of Equity, Diversity, Inclusion & Belonging, I wish to express my deepest admiration and appreciation.

@@

To my parents, your \emph{hijito} hopes that this work honors your lifelong commitment to the pursuit of knowledge and a better world—a devotion only exceeded by the love and support you have shared with me every day of my life.

@@

To John, my partner, words cannot express how much it has meant to me to call you that. Every step on this journey, love, has been easier with you at my side, with your hand in mine.

@@

\emph{To the steps we have taken already, and to the many we have ahead.}

>>>>

First and foremost, I would like to thank the members of my dissertation committee, Stéphane Verguet, Jessica Cohen, Karestan Koenen, and Margaret McConnell. I thank the committee chair, Stéphane Verguet, for his service, and for his extraordinary mentorship as my academic advisor. Throughout the last four years, he has consistently offered unwavering support and insightful guidance throughout my academic journey, and in the moments where I faltered or doubted myself most, his confidence and cheer always helped see me through.

Jessica Cohen has been a constant source of joy and inspiration throughout my studies, and her teaching and scholarship have set a standard to aspire to. In addition to being a brilliant scholar, she is an incredible mentor and instructor, bringing genuine joy to the rigorous study of challenging questions. I feel extremely grateful to have learned so much from her example and from her enthusiasm for econometrics and public health.

@@

I am also deeply grateful to my coauthors and reviewers, whose thoughtful reflections strengthened every page. I thank Shekhar Saxena for his guiding leadership in global mental health and for welcoming me into the field when I first arrived at Harvard, and I thank Christopher Golden for welcoming me into his lab and providing every opportunity, guidance, and support a student could wish for throughout the development of the second paper of this thesis.

@@

The community in the Department of Global Health and Population has been a treasured home away from home, and I am grateful to so many of its members for the opportunity to learn at their side. My deep appreciation goes to David Bloom, David Canning, Margaret Kruk, and Vikram Patel, whose classes and coffee chats all have shaped my interests and the course of my career. I also owe a debt of gratitude to Barbara Heil and Allison Conary for all the mountains they have moved and the uphill struggles they have seen me soundly through.

I also would like to thank the many other individuals at the School of Public Health who have made the last four years so full of warm memories. I am thankful to Claudette Agustin, Matthew Boccuzzi, Jarvis Chen, and Bruce Villineau for all the care and welcome they have shared with me and with each cohort of students in the Population Health Sciences program.

@@

Across the Charles River are many dear colleagues who have dedicated themselves to ensuring that Harvard feels like home to all its students. To Jackie Yun and Janet Daniels at the GSAS Student Center, and to Sheila Thomas, Katie Saibara, and Karina Gonzalez Herrera of the GSAS Office of Equity, Diversity, Inclusion & Belonging, I wish to express my deepest admiration and appreciation.

@@

To my parents, your \emph{hijito} hopes that this work honors your lifelong commitment to the pursuit of knowledge and of a better world—a devotion only exceeded by the love and support you have shared with me every day of my life.

@@

To John, my partner, words cannot express how much it has meant to me to call you that. Every step on this journey has been easier with you at my side, with your hand in mine.

@@

\emph{To the steps we have taken already, love, and to the many we have ahead.}

### 2023-03-11 intro.Rmd
Revisions throughout and final version for External Reader

For much of the global population, however, attaining this state of mental health is an enduring challenge, with over one billion people worldwide living with a mental or addictive disorder.[@rehmGlobalBurdenDisease2019] Mental disorders are a leading cause of disability,[@charlsonExcessMortalityMental2015; @rehmGlobalBurdenDisease2019] with the Global Burden of Disease (GBD) 2019 study estimating that 4.9% of disability-adjusted life years (DALYs) and 14.6% of years lived with disability (YLDs) are attributable to mental disorders.[@vosGlobalBurden3692020a] Mental disorders are closely linked to physical health, and can be a significant risk factor for premature mortality.[@walkerMortalityMentalDisorders2015] Among patients living with the most severe mental illnesses, premature mortality may reduce life expectancy by as much as 10 to 20 years.[@demooijDyingTooSoon2019]

At all levels of sociodemographic development, this burden of morbidity and mortality is rising.[@patelLancetCommissionGlobal2018] Yet despite this global scale and despite a growing evidence base for impactful, cost-effective interventions that may help,[@patelMentalNeurologicalSubstance2016] the profound impact of mental disorders has been met by a profound treatment gap.[@demyttenaerePrevalenceSeverityUnmet2004] The majority of people living with a mental disorder do not have access to adequate care (Figure \ref{figure_service_gap}),[@bassUnitedNationsGeneral2012a; @demyttenaerePrevalenceSeverityUnmet2004] and many face stigma,[@alonsoPerceivedStigmaIndividuals2009; @angermeyerStigmaMentalIllness2003] discrimination,[@stuartMentalIllnessEmployment2006; @thornicroftDiscriminationHealthCare2007] and abuse.[@humanrightswatchDeathSentenceAbuses2012; @mfoafo-mcarthyHumanRightsViolations2014; @rahmanMentalDistressHuman2021]

@@

```{r figure_service_gap, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 11, fig.height = 6.5, fig.align='center', out.width = '97%',  fig.scap="Countries where fewer than 10 psychiastrists per 100,000 capita were working in the health sector, 2013-2017.", out.extra='',  fig.cap="\\label{figure_all_diseases}Countries where fewer than 10 psychiastrists per 100,000 capita were working in the health sector between 2013-2017, average. Data from the Globa Health Observator (GHO) of the World Health Organization (WHO)."}

@@

In addition to the long-standing challenges of these unmet needs, new and emerging dangers such as climate change and COVID-19 reveal additional stressors that threaten to exacerbate this burden. Global climate change—an accelerating existential threat to planetary health that the United Nations considers the “defining crisis of our time”—is already manifesting profound impacts on mental health.[@worldhealthorganizationMentalHealthClimate2022] The impacts are shaped both by the disastrous upheaval caused by worsening extreme weather events and by the incremental change rising temperatures and sea levels have wrought on livelihoods and means of living—altering the availability and access to natural resources, land use, infrastructure, and community.[@atwoliMentalHealthClimate2022; @hayesClimateChangeMental2018] This essential connections between mental health and human wellbeing have been further underscored by the devastating impacts of the COVID-19 pandemic, which continues to deal psychological and psychiatric harm to patients, health care workers, and the public overall.[@vindegaardCOVID19PandemicMental2020]

The extant and evolving challenges of global mental health, in turn, entail jeopardy to health systems and to the three intrinsic goals of improving health, responsiveness, and financial protection. According to the WHO’s Mental Health Atlas report from 2020, mental health system governance, capacity, and integration into routine service delivery is fundamentally lacking across nations of all levels of economic development, with only 15% of countries having functional integration of mental health into primary health care.[@worldhealthorganizationMentalHealthAtlas2021] This lack of service delivery echoes a lack of basic epidemiological data,[@steinCrossnationalEpidemiologySocial2017] constraining researching and frustrating efforts to respond to mental health needs. Among nations, the Atlas report highlights widening disparities in health system indicators, with worsened gaps in service availability and public health expenditure for mental health between low- and middle-income countries and high-income peers. With respect to health spending in many countries, mental health care services have relied primarily or exclusively on out-of-pocket payments,[@dixonFinancingMentalHealth2006] resulting in a greater risk of catastrophic health expenditure among households with patients living with a mental disorder. These direct costs, when coupled with the indirect costs of diminished due to disability, may exacerbate conditions of poverty, [@patelPovertyCommonMental2003] which, in turn, may lead to worsening mental health[@lundPovertyCommonMental2010]—exacerbating a vicious cycle that harms households and nations alike.

A better understanding of the evolving burden and dynamics of global mental health may thus have broad consequence to strengthening of health system responses to mental disorders and improving health outcomes for millions of individuals worldwide. Contributing to such an understanding has motivated this dissertation, which takes as its focus different quantitative aspects of the global burden of mental illness—how we measure and evaluate it, how it may be changing, and how novel data sources may help in characterizing it, particularly in the absence of robust, high-frequency epidemiological surveillance. The papers presented in the subsequent three chapters echo the interdisciplinary nature of global mental health, drawing on a broad range of quantitative methods and applying analytic tools and insights from epidemiology, health economics, and econometrics.

In Chapter 1, we investigated the share of the global burden of disease attributable to mental disorders and its associated economic value. We reviewed recent efforts to estimate both the epidemiologic and economic burden of mental disorders, and we propose a composite approach to estimation to capture a broader share of attributable morbidity and mortality.[@bloomGlobalEconomicBurden2011; @charlsonExcessMortalityMental2015; @vigoEstimatingTrueGlobal2016; @vosGlobalBurden3692020a; @walkerMortalityMentalDisorders2015; @whitefordGlobalBurdenDisease2013] Applying this approach using the most recently available GBD data, we estimated that the global burden of mental disorders is likely much higher than previously estimated, encompassing 16% of disability-adjusted life years in the year 2019. Using value of statistical life approaches, we estimated that the economic value of this mental health burden would exceed 4.7 trillion USD.

In Chapter 2, we investigated how health system visits for mental disorders in Madagascar may be shifting in respond to changes in climate. Madagascar has little to no mental health surveillance data and is among the nations most vulnerable to climate change. To address these gaps, we studied the impact of three climate change exposures—temperature, soil moisture, and duration of cyclones and tropical storms—by leveraging meteorologic, geospatial, and health system data reported by 3,413 facilities in Madagascar from 2010 to 2020. We found a statistically significant reduction in monthly cases reported associated with temperature, primarily localized to facilities in the central highlands, where cooler temperatures are more prevalent. We further found that higher soil moisture may be associated with higher case counts in the future, after a three month lag; this increase was primarily found in the eastern lowlands and the northern dry deciduous forests, consistent with the geographic concentration of flood risk in Madagascar. Our findings point to suggest potential heterogeneity in how changing climate conditions will impact mental health needs and system responses.

In Chapter 3, we investigated how internet search data for health symptoms might provide high frequency insight into the state of population mental health during the COVID-19 pandemic. Using search data from U.S. counties, states, and international trends, we explored how announcements of COVID-19 vaccine safety and efficacy data in November 2020 impacted Google search trends for mental health symptoms. We found evidence of statistically significant immediate and sustained declines in searches for anxiety, depression, and major depressive disorder throughout the 40 days following these announcements. The findings underscore the importance of timely and transparent communication and illustrate the potential application of high frequency internet search data for population mental health surveillance.

>>>>

For much of the global population, however, attaining this state of mental health is an enduring challenge, with over one billion people worldwide living with a mental or addictive disorder.[@rehmGlobalBurdenDisease2019] Mental disorders are a leading cause of disability,[@charlsonExcessMortalityMental2015; @rehmGlobalBurdenDisease2019] with the Global Burden of Disease (GBD) study estimating that 4.9% of disability-adjusted life years (DALYs) and 14.6% of years lived with disability (YLDs) were attributable to mental disorders in 2019.[@vosGlobalBurden3692020a] Mental disorders are also closely linked to physical health and can be a significant risk factor for premature mortality.[@walkerMortalityMentalDisorders2015] Among patients living with the most severe mental illnesses, premature mortality may reduce life expectancy by as much as 10 to 20 years.[@demooijDyingTooSoon2019]

At all levels of sociodemographic development, this burden is rising.[@patelLancetCommissionGlobal2018] Yet despite this global scale and despite a growing evidence base for impactful, cost-effective interventions that may help,[@patelMentalNeurologicalSubstance2016] the profound impact of mental disorders has been met by a profound treatment gap.[@demyttenaerePrevalenceSeverityUnmet2004] A majority of the world's population lives in countries with fewer than 10 psychiatrists for every 100,000 people (Figure \ref{figure_service_gap}). Most of those living with a mental disorder do not have access to adequate care,[@bassUnitedNationsGeneral2012a; @demyttenaerePrevalenceSeverityUnmet2004] and many face stigma,[@alonsoPerceivedStigmaIndividuals2009; @angermeyerStigmaMentalIllness2003] discrimination,[@stuartMentalIllnessEmployment2006; @thornicroftDiscriminationHealthCare2007] and abuse.[@humanrightswatchDeathSentenceAbuses2012; @mfoafo-mcarthyHumanRightsViolations2014; @rahmanMentalDistressHuman2021]

@@

```{r figure_service_gap, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 11, fig.height = 6.5, fig.align='center', out.width = '97%', fig.scap="Countries where fewer than 10 psychiatrists per 100,000 capita were working in the health sector, 2013-2017.", out.extra='', fig.cap="\\label{figure_service_gap}Countries where fewer than 10 psychiatrists per 100,000 capita were working in the health sector between 2013-2017, average. Data from the Global Health Observatory (GHO) of the World Health Organization (WHO)."}

@@

In addition to the long-standing challenges of these unmet needs, new and emerging dangers such as climate change and COVID-19 reveal additional stressors that threaten to exacerbate this burden. Global climate change—an accelerating existential threat to planetary health that the United Nations considers the “defining crisis of our time”—is already manifesting profound impacts on mental health.[@worldhealthorganizationMentalHealthClimate2022] These impacts are shaped both by the disastrous upheaval caused by worsening extreme weather events and by the incremental change rising temperatures and sea levels have wrought on livelihoods and means of living—altering the availability and access to natural resources, land use, infrastructure, and community.[@atwoliMentalHealthClimate2022; @hayesClimateChangeMental2018] The essential connections between mental health and human well-being have been further underscored by the devastating impacts of the COVID-19 pandemic, which continues to deal psychological and psychiatric harm to patients, health care workers, and the public overall.[@vindegaardCOVID19PandemicMental2020]

The extant and evolving challenges of global mental health, in turn, entail jeopardy to health systems and to the three intrinsic goals of improving health, responsiveness, and financial protection.[@robertsGettingHealthReform2008] According to the WHO’s Mental Health Atlas report from 2020, mental health system governance, capacity, and integration into routine service delivery is fundamentally lacking across nations of all levels of economic development, with only 15% of countries having functional integration of mental health into primary health care.[@worldhealthorganizationMentalHealthAtlas2021] This lack of service delivery echoes a lack of basic epidemiological data,[@steinCrossnationalEpidemiologySocial2017] constraining research and frustrating efforts to respond to mental health needs. Among nations, the Atlas report highlights widening disparities in health system indicators, with worsened gaps in service availability and public health expenditure for mental health between low- and middle-income countries and high-income peers. With respect to health spending, mental health care services in many countries have traditionally relied primarily or exclusively on out-of-pocket payments,[@dixonFinancingMentalHealth2006] resulting in a greater risk of catastrophic health expenditure among households with patients living with a mental disorder. These direct costs, when coupled with the indirect costs of diminished earnings due to disability, may exacerbate conditions of poverty, [@patelPovertyCommonMental2003] which, in turn, may lead to worsening mental health[@lundPovertyCommonMental2010]—feeding a vicious cycle that harms households and nations alike.

A better understanding of the evolving burden and dynamics of global mental health may thus have broad consequences to strengthening health system responses to mental disorders and improving health outcomes for millions of individuals worldwide. Contributing to such an understanding has motivated this dissertation, which takes as its focus different quantitative aspects of the global burden of mental illness—how we measure and evaluate it, how it may be changing, and how novel data sources may help in characterizing it, particularly in the absence of robust, high-frequency epidemiological surveillance. The papers presented in the subsequent three chapters echo the interdisciplinary nature of global mental health, drawing on a broad range of quantitative methods and applying analytic tools and insights from epidemiology, health economics, and econometrics.

In Chapter 1, we investigated the share of the global burden of disease attributable to mental disorders and its associated economic value. We reviewed recent efforts to estimate both the epidemiologic and economic burden of mental disorders, and we propose a composite approach to estimation to capture a broader share of attributable morbidity and mortality.[@bloomGlobalEconomicBurden2011; @charlsonExcessMortalityMental2015; @vigoEstimatingTrueGlobal2016; @vosGlobalBurden3692020a; @walkerMortalityMentalDisorders2015; @whitefordGlobalBurdenDisease2013] Applying this approach using the most recently available GBD data, we estimated that the global burden of mental disorders is likely much higher than previously estimated, encompassing 16% of disability-adjusted life years in 2019. Using value of statistical life approaches, we estimated that the economic value of this burden would exceed 4.7 trillion USD.

In Chapter 2, we investigated how health system visits for mental disorders in Madagascar may be shifting in respond to changes in climate. Madagascar has little to no mental health surveillance data and is among the nations most vulnerable to climate change. To address these gaps, we studied the impact of three climate change exposures—temperature, soil moisture, and duration of cyclones and tropical storms—by leveraging meteorologic, geospatial, and health system data reported by 3,413 facilities from 2010 to 2020. We found a statistically significant reduction in monthly mental health related visits associated with each degree Celcius increase in temperature, primarily localized to facilities in the central highlands, where cooler temperatures are more prevalent. We further found that higher soil moisture may be associated with more mental health visits in the future, after a three month lag; this increase was primarily found in the eastern lowlands and the northern dry deciduous forests, consistent with the geographic concentration of flood risk in Madagascar. Our findings point to potential heterogeneity in how changing climate conditions will impact mental health needs and system responses.

In Chapter 3, I investigated how internet search data for health symptoms might provide high frequency insight into the state of population mental health during the COVID-19 pandemic. Using search data from U.S. counties, states, and international trends, I explored how announcements of COVID-19 vaccine safety and efficacy data in November 2020 impacted Google search trends for mental health symptoms. I found evidence of statistically significant immediate and sustained declines in searches for anxiety, depression, and major depressive disorder throughout the 40 days following these announcements. The findings underscore the importance of timely and transparent communication and illustrate the potential application of high frequency internet search data for population mental health surveillance.

### 2023-03-11 paper1.Rmd
Revisions throughout and final version for External Reader

# Loading libraries
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RColorBrewer)
library(sp)
library(broom)
library(sf)
library(scales)
library(viridis)
library(data.table)
library(ggpubr)
library(lemon)
library(rnaturalearth)
library(rnaturalearthdata)
library(sysfonts)
library(showtext)
font_add_google("Source Sans Pro")
library(gridExtra)
showtext_auto() 
showtext_opts(dpi = 300)  
library(rnaturalearthhires) # install.packages("rnaturalearthhires", repos = "http://packages.ropensci.org", type = "source")

@@

results_sentence1 <- paste0(
  "Under GBD 2019, over ",
  results1,
  " million DALYs were attributed to mental disorders, or roughly ",
  results2,
  "\\% of the global burden. After including alcohol and drug use, neurological disorders, chronic pain, suicide, and self-harm, the share due to mental disorders rose to ",
  results3,
  "\\% of global DALYs (approximately ",
  results4,
  " million DALYs). Under the composite approach, an additional  ",
  abstract1-results4,
  " million DALYs were attributed to mental disorders, encompassing, in total, over ",
  abstract2,
  "\\% of global DALYs (Figure 1). Under all three methods, the burden of mental disorders (in DALYs) exhibited a country-income gradient, with mental disorders comprising ",
  ifelse(results5/results6 > 2.4, "over twice", ifelse(results5/results6> 2, "roughly twice", "approximately twice")),
  " the burden of disease in high-income countries compared to low-income countries.")

@@

\emph{Daniel Arias, Shekhar Saxena, Stéphane Verguet}

@@

A version of this paper was accepted for publication by \emph{eClinicalMedicine} on September 9, 2022 and was published in Volume 54, Issue 101675 in December 2022.

@@

The most recent estimates of morbidity and mortality due to mental disorders come from the Global Burden of Disease (GBD) 2019 study.[@vosGlobalBurden3692020a] The GBD study provides disease burden estimates using DALYs, years of life lost (YLLs), and years lived with disability (YLDs), which are then aggregated within a hierarchical grouping scheme that classifies causes of disability and death at different levels of mutually exclusive and completely exhaustive categories. (Mental disorders are a Level 2 condition, nested under NCDs; see Table S1.3.)

While GBD remains the gold standard for global epidemiologic estimation, the nature of the GBD scheme—in particular, the rationale for grouping certain conditions under mental disorders or not—has been the subject of debate in the literature.[@atunChallengesEstimatingTrue2016; @vigoEstimatingTrueGlobal2016; @whitefordChallengesEstimatingTrue2016] In particular, work by Vigo et al. (2016) published in The Lancet Psychiatry argues for an expanded classification of mental disorders under the GBD classification scheme to account for underestimation of the burden of mental disorders.[@vigoEstimatingTrueGlobal2016] The authors attribute this underestimation to five main causes: 1) the distinction drawn between mental and neurological diseases; 2) the categorization of self-harm and suicide under injuries; 3) the classification of all chronic pain and somatoform disorders under musculoskeletal disorders; 4) the exclusion of personality disorders; and 5) the exclusion of premature mortality due to mental disorders. Using data from the 2013 GBD study, Vigo and colleagues re-allocated the entire burden of dementias, epilepsy, migraine, tension-type headache, and self-harm to mental disorders. In addition, a third of the burden of musculoskeletal disorders without anatomical correlate (i.e., somatoform disorders with prominent pain) was attributed to mental disorders.[@vigoEstimatingTrueGlobal2016] This reallocation attributed 13% of DALYs to mental disorders, a 6 percentage point increase from the GBD estimate of 7%.

In this paper, we attempt to revisit the estimation of the global burden of mental disorders and of its associated economic value. Our aim is to characterize potential underestimation of the burden of mental disorders and to quantify the economic value of this burden under different estimation approaches. Specifically, we expand on Vigo et al.[@vigoEstimatingTrueGlobal2016] by capturing premature mortality due to mental disorders using pooled risk ratios of mortality from a systematic review of mental disorders[@walkerMortalityMentalDisorders2015] to determine the population attributable fraction (PAF) of premature mortality. Inclusion of premature mortality through the PAF presents a novel composite approach that can more broadly capture attributable morbidity and mortality. Using this approach on GBD 2019 estimates, we then apply monetary values to DALYs to reach estimates of the global economic value of the mental burden of disease using a value of a statistical life (VSL) approach similar to that of Bloom and colleagues,[@bloomGlobalEconomicBurden2011] which attempts to capture a population’s willingness to pay to reduce morbidity and mortality associated with illness. The VSL approach—in contrast alternatives such as cost of illness and value of lost output approaches—includes an economic valuation of mortality risk reductions in monetary terms, and thus enables comparison across sectors (beyond the sole health sector) which can motivate decision-making toward ameliorating welfare and societal mental health. Our findings suggest that both the epidemiological and economic burden of mental disorders could be larger than previously estimated, and that underestimation would be larger among regions where premature mortality due to mental disorders is greater.

@@

In addition to our base-case economic valuation, we further report our VSL estimates using three times GDP per capita as the value of a DALY. (Alternative valuations using values of $1,000 and $5,000, as well as purchasing power parity (PPP)-adjusted GDP per capita, are reported in Tables S1.5 and S1.6 of Appendix 1)

@@

The research draws exclusively on secondary country-level data reported at the national or subnational level. As such, it does not involve data collection, experimentation, or investigation concerning human subjects. The Institutional Review Board (IRB) of the Harvard T.H. Chan School of Public Health determined that the study was not human subjects research, and that additional review was not required (protocol number: IRB20-1946, determined on November 13, 2020).

### Role of funding source

This study received no funding. All authors (DA, SV, and SS) had access to the data and shared in the decision to submit this article for publication.

@@

```{r figure_dalys_nation, include = TRUE, echo = FALSE, warnings = FALSE, fig.scap="DALYs attributable to mental disorders in 2019, by estimation approach, as a share of all DALYs.", out.extra='', fig.cap=paste0("\\label{figure_dalys_nation}DALYs attributable to mental disorders in 2019, by estimation approach, as a share of all DALYs. ATG: Antigua and Barbuda; DALYs: disability-adjusted life years; FSM: Federated States of Micronesia; LCA: Saint Lucia; TLS: East Timor; TTO: Trinidad and Tobago; VCT: Saint Vincent and the Grenadines."), fig.width = 13.7, fig.height = 7.5, fig.align='center', out.width = '98%'}

@@

```{r figure_gbd_adjusted, include = TRUE, echo=FALSE, warnings = FALSE, fig.scap="DALYs, YLDs, YLLs, and deaths attributable to mental disorders in 2019, by estimation approach, per 100,000 population.", out.extra='', fig.cap=paste0("\\label{figure_gbd_adjusted}DALYs, YLDs, YLLs, and deaths attributable to mental disorders in 2019, by estimation approach, per 100,000 population. Values are aggregated by GBD region. DALYs: Disability-adjusted life years; YLDs: years lived with disability; YLLs: years of life lost; GBD: Global Burden of Disease."), fig.width = 11, fig.height = 11, fig.align='center', out.width = '97%'}

@@

```{r figure_cost_by_region, fig.width = 8, fig.height = 6, fig.align='center', out.width = '95%',  fig.scap="Economic burden of mental disorders, as a percent of GDP.", out.extra='', fig.cap="\\label{figure_cost_by_region}Economic burden of mental disorders, as a percent of GDP. The economic value is determined by using GDP per capita (USD 2019) as the value of a DALY. Values are aggregated by GBD region. GDP: gross domestic product; USD: United States dollar; DALY: disability-adjusted life year; GBD: Global Burden of Disease.", echo = FALSE, include = TRUE}

@@

All authors contributed to study conception, methodology, and interpretation. DA oversaw data acquisition, programming, formal analysis, visualization, the first draft of the manuscript. All authors contributed to critical revision of the manuscript, with responses to reviewers and subsequent revisions led by DA. All authors had access to and verified all the data and accept responsibility for the decision to submit for publication.

>>>>

@@

results_sentence1 <- paste0(
 "Under GBD 2019, over ",
 results1,
 " million DALYs were attributed to mental disorders, or roughly ",
 results2,
 "\\% of the global burden. After including alcohol and drug use, neurological disorders, chronic pain, suicide, and self-harm, the share due to mental disorders rose to ",
 results3,
 "\\% of global DALYs (approximately ",
 results4,
 " million DALYs). Under the composite approach, an additional ",
 abstract1-results4,
 " million DALYs were attributed to mental disorders, encompassing, in total, over ",
 abstract2,
 "\\% of global DALYs (Figure 2). Under all three methods, the burden of mental disorders (in DALYs) exhibited a country-income gradient, with mental disorders comprising ",
 ifelse(results5/results6 > 2.4, "over twice", ifelse(results5/results6> 2, "roughly twice", "approximately twice")),
 " the burden of disease in high-income countries compared to low-income countries.")

@@

\emph{Daniel Arias, Shekhar Saxena, and Stéphane Verguet}

@@

An earlier version of this paper was published by \emph{eClinicalMedicine} in Volume 54, Issue 101675 in December 2022.

@@

The most recent estimates of morbidity and mortality due to mental disorders come from the Global Burden of Disease (GBD) 2019 study.[@vosGlobalBurden3692020a] The GBD study provides disease burden estimates using DALYs, years of life lost (YLLs), and years lived with disability (YLDs), which are then aggregated within a hierarchical grouping scheme that classifies causes of disability and death at different levels of mutually exclusive and completely exhaustive categories. (Mental disorders are a Level 2 condition, nested under noncommunicable diseases (NCDs); see Table S1.3.)

While GBD remains the gold standard for global epidemiologic estimation, the nature of the GBD scheme—in particular, the rationale for grouping certain conditions under mental disorders or not—has been the subject of debate in the literature.[@atunChallengesEstimatingTrue2016; @vigoEstimatingTrueGlobal2016; @whitefordChallengesEstimatingTrue2016] In particular, work by Vigo et al. (2016) published in \emph{The Lancet Psychiatry} argues for an expanded classification of mental disorders under the GBD classification scheme to account for underestimation of the burden of mental disorders.[@vigoEstimatingTrueGlobal2016] The authors attribute this underestimation to five main causes: 1) the distinction drawn between mental and neurological diseases; 2) the categorization of self-harm and suicide under injuries; 3) the classification of all chronic pain and somatoform disorders under musculoskeletal disorders; 4) the exclusion of personality disorders; and 5) the exclusion of premature mortality due to mental disorders. Using data from the 2013 GBD study, Vigo and colleagues re-allocated the entire burden of dementias, epilepsy, migraine, tension-type headache, and self-harm to mental disorders. In addition, a third of the burden of musculoskeletal disorders without anatomical correlate (i.e., somatoform disorders with prominent pain) was attributed to mental disorders.[@vigoEstimatingTrueGlobal2016] This reallocation attributed 13% of DALYs to mental disorders, a 6 percentage point increase from the GBD estimate of 7%.

In this paper, we attempt to revisit the estimation of the global burden of mental disorders and of its associated economic value. Our aim is to characterize potential underestimation of the burden of mental disorders and to quantify the economic value of this burden under different estimation approaches. Specifically, we expand on Vigo et al.[@vigoEstimatingTrueGlobal2016] by capturing premature mortality due to mental disorders using pooled risk ratios of mortality from a systematic review of mental disorders[@walkerMortalityMentalDisorders2015] to determine the population attributable fraction (PAF) of premature mortality. Inclusion of premature mortality through the PAF presents a novel composite approach that can more broadly capture attributable morbidity and mortality. Using this approach on GBD 2019 estimates, we then apply monetary values to DALYs to reach estimates of the global economic value of the mental burden of disease using a value of a statistical life (VSL) approach similar to that of Bloom and colleagues,[@bloomGlobalEconomicBurden2011] which attempts to capture a population’s willingness to pay to reduce morbidity and mortality associated with illness. The VSL approach—in contrast to alternatives such as cost of illness and value of lost output approaches—includes an economic valuation of mortality risk reductions in monetary terms, and thus enables comparison across sectors (beyond the sole health sector) which can motivate decision-making toward ameliorating welfare and societal mental health. Our findings suggest that both the epidemiological and economic burden of mental disorders could be larger than previously estimated, and that underestimation would be larger among regions where premature mortality due to mental disorders is greater.

@@

In addition to our base-case economic valuation, we further report our VSL estimates using three times GDP per capita as the value of a DALY. (Alternative valuations using values of $1,000 and $5,000, as well as purchasing power parity (PPP)-adjusted GDP per capita, are reported in Tables S1.5 and S1.6 of Appendix 1.)

@@

The Institutional Review Board (IRB) of the Harvard T.H. Chan School of Public Health determined that this study was not human subjects research and that additional review was not required (protocol number: IRB20-1946, determined on November 13, 2020).

@@

```{r figure_dalys_nation, include = TRUE, echo = FALSE, warnings = FALSE, fig.scap="Disability-adjusted life years (DALYs) attributable to mental disorders in 2019, by estimation approach, as a share of all DALYs.", out.extra='', fig.cap=paste0("\\label{figure_dalys_nation}Disability-adjusted life years (DALYs) attributable to mental disorders in 2019, by estimation approach, as a share of all DALYs. ATG: Antigua and Barbuda. FSM: Federated States of Micronesia. LCA: Saint Lucia. TLS: East Timor. TTO: Trinidad and Tobago. VCT: Saint Vincent and the Grenadines."), fig.width = 13.7, fig.height = 7.5, fig.align='center', out.width = '98%'}

@@

```{r figure_gbd_adjusted, include = TRUE, echo=FALSE, warnings = FALSE, fig.scap="DALYs, YLDs, YLLs, and deaths attributable to mental disorders in 2019, by estimation approach, per 100,000 population.", out.extra='', fig.cap=paste0("\\label{figure_gbd_adjusted}DALYs, YLDs, YLLs, and deaths attributable to mental disorders in 2019, by estimation approach, per 100,000 population. Values are aggregated by GBD region. DALYs: Disability-adjusted life years. YLDs: years lived with disability. YLLs: years of life lost. GBD: Global Burden of Disease."), fig.width = 11, fig.height = 11, fig.align='center', out.width = '97%'}

@@

```{r figure_cost_by_region, fig.width = 8, fig.height = 6, fig.align='center', out.width = '95%', fig.scap="Economic burden of mental disorders, as a percent of GDP.", out.extra='', fig.cap="\\label{figure_cost_by_region}Economic burden of mental disorders, as a percent of GDP. The economic value is determined by using GDP per capita (USD 2019) as the value of a DALY. Values are aggregated by GBD region. GDP: gross domestic product. USD: United States dollar. DALY: disability-adjusted life year. GBD: Global Burden of Disease.", echo = FALSE, include = TRUE}

@@

All authors contributed to study conception, methodology, and interpretation. DA oversaw data acquisition, programming, formal analysis, visualization, and the first draft of the manuscript. All authors contributed to critical revision of the manuscript, with responses to reviewers and subsequent revisions led by DA.

### Funding statement

This study received no funding.
